### 2024-05-03 scripts/11-proof-of-concept.qmd

@@

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
theme_set(theme_bw())
library(tmfast)

@@

This gives us three sets of journals: A, B (0-5), and C-L.  In the LDA model, these journal sets correspond to $\mathbf{\alpha}$, the parameter for the "prior" Dirichlet distribution $\Theta(\mathbf{\alpha})$. The `tmfast` package provides a helper function `peak_alpha()` for constructing $\mathbf{\alpha}$.

@@

Each paper has a simplex $\theta \sim \Theta(\mathbf{\alpha})$, representing a distribution over topics. We draw `r N` papers from each journal.

@@

head(theta)

@@

```{r}
## Fit topic model ----
fitted = tmfast(corpus, c(10, 15, 20, 25, 30))
```

@@

```{r}
## Topics by journal ----
add_journal = function(x) {
    mutate(x, journal = str_extract(document, '^[A-Z]*[0-9]?'))
}

tidy(fitted, k = 20, matrix = 'gamma') |> 
    add_journal() |> 
    ggplot(aes(document, topic, fill = gamma)) +
    geom_tile() +
    facet_wrap(vars(journal), scales = 'free', 
               ncol = 7) +
    scale_x_discrete(guide = 'none')
```

@@

```{r}
## Topic model clearly detects the focal topic in A, increase in B over time, minimal presence in other journals
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red')
```

While the journals are largely distinguishable on the other topics, except for the unusually low value for A. 
```{r}
## With no pattern for the other topics
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V02') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red')
```

@@

```{r}
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01', 
           gamma > .1) |> 
    add_journal() |> 
    count(journal) |> 
    mutate(rate = n / N)
```

@@

As a limitation, we note that, for $\theta > 0.05$ or so, the estimated topic probabilities are below the true values. 
```{r}
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    inner_join(theta_df, by = c('document' = 'doc', 'topic')) |> 
    ggplot(aes(theta, gamma, color = journal)) +
    geom_point() +
    stat_function(fun = identity, 
                  inherit.aes = FALSE, color = 'black') +
    scale_color_viridis_d()
```

>>>>

---
format:
  - html
  - pdf: 
      mathspec: true
---

@@

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
library(ggforce)
theme_set(theme_bw())
library(tmfast)

@@

This gives us three sets of journals: A, B (0-5), and C-L.  In the LDA model, these journal sets correspond to $\boldsymbol{\alpha}$, the parameter for the "prior" Dirichlet distribution $\Theta(\boldsymbol{\alpha})$. The `tmfast` package provides a helper function `peak_alpha()` for constructing $\boldsymbol{\alpha}$.

@@

Each paper has a simplex $\theta \sim \Theta(\boldsymbol{\alpha})$, representing a distribution over topics. We draw `r N` papers from each journal.

@@

theta[1:5, 1:5]

@@

```{r}
## Fit topic model ----
tic()
fitted = tmfast(corpus, c(10, 15, 20, 25, 30))
toc()
```

@@

```{r}
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Fitted topic-document distributions for the proof-of-concept simulation."
## Topics by journal ----
add_journal = function(x) {
    mutate(x, journal = str_extract(document, '^[A-Z]*[0-9]?'))
}

tidy(fitted, k = 20, matrix = 'gamma') |> 
    add_journal() |> 
    ggplot(aes(document, topic, fill = gamma)) +
    geom_raster() +
    facet_wrap(vars(journal), scales = 'free_x', 
               ncol = 7) +
    scale_y_discrete(breaks = c('V01', 'V10', 'V20')) +
    scale_x_discrete(breaks = 'none')
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, the topic model clearly detects the marginal topic in journal A, increase in journal B over time, and minimal presence in other journals"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red')
```

While the journals are largely not distinguishable from each other on the other topics, except for the unusually low value for A.

```{r}
#| fig-cap: "In the proof-of-concept simulation, the topic model does not produce a false negative pattern with the mainstream topics"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V02') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    geom_sina() +
    stat_summary(color = 'red')
```

@@

```{r}
#| tbl-cap: "In the proof-of-concept simulation, almost all articles in journal A contain the marginal topic; the rate in B increases over time; and the marginal topic does not appear in the other journals"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01', 
           gamma > .1) |> 
    add_journal() |> 
    count(journal, .drop = FALSE) |> 
    mutate(rate = n / N) |> 
    knitr::kable()
```

@@

As a limitation, we note that, for $\theta > 0.05$ or so, the estimated topic probabilities are below the true values.

```{r}
#| fig-cap: "In the proof-of-concept simulation, without normalization the estimated topic-document probabilities (gamma) are well below the true probabilities (theta) for approx. $\\theta > 0.05$"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    inner_join(theta_df, by = c('document' = 'doc', 'topic')) |> 
    ggplot(aes(theta, gamma, color = journal)) +
    geom_point() +
    stat_function(fun = identity, 
                  inherit.aes = FALSE, color = 'black') +
    scale_color_viridis_d(guide = 'none')
```
